`Total-evidence' Bayesian estimation of phylogeny, divergence times and fossil ages

### 2015-10-29 TotalEvidenceTalk.tex
homo naledi

\usepackage{tabularx}
\usepackage[]{algorithm2e}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
\bibliography{fossilDating.bib}

\newcommand{\Mstrict}{{M1}}
\newcommand{\Mrelaxed}{{M8}}

@@

\begin{frame}
\frametitle{The fossilized birth-death process}

@@

\begin{frame}
\begin{figure}
\includegraphics[width=0.6\textwidth]{priors.pdf}
\label{Fig:Prior}
\end{figure}

Density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.
\end{frame}

>>>>

\usepackage{tabularx}
\usepackage{tikz}
\usepackage[]{algorithm2e}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{subfigure}
\usepackage[style=authoryear,backend=biber,mincitenames=1,maxcitenames=2, uniquelist=false]{biblatex}
\bibliography{fossilDating.bib}

\newcommand{\Mstrict}{{M1}}
\newcommand{\Mrelaxed}{{M8}}
% turnover = birth / death
\newcommand{\turnover}{\nu}

% sampling proportion = \psi / \psi + \mu
\newcommand{\fosp}{s}
\newcommand{\sampleprop}{sampling proportion}

\definecolor{newblue}{rgb}{0.3, 0.5, 0.9}
\definecolor{newblue1}{rgb}{0.3, 0.3, 0.7}
\definecolor{newred}{RGB}{200,0,0}

@@

\begin{frame}{The problem}

\begin{itemize}

\item We have 
\begin{itemize}
  \item \normalsize{{\bf molecular data} of extant species,} 
  \item \normalsize{{\bf morphological data} of extant and fossil species and }
  \item \normalsize{{\bf Fossil ages} (or geological time intervals).}
\end{itemize} 
\item We want to utilise all this data to learn about {\bf evolutionary history} and {\bf macroevolutionary processes}.
\item {\bf Bayesian statistical inference} is our preferred method of learning about the world.

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Node dating}

\begin{itemize}
\item {\bf Divergence time dating is tricky :) }
\item {\bf A common practice is node dating}. One or more divergences in a phylogeny are assigned ages by identifying fossils of known geological age that correspond* to each divergence. 
\item This method has a few drawbacks:
\begin{itemize}
\item Often only the oldest fossil in a calibrated clade is used,
\item Calibration densities are often {\it ad hoc},
\item Difficult* to combine calibrations with tree priors,
\item Computational complexity of some methods mean that only a few nodes can be calibrated (Heled \& Drummond 2012).
\end{itemize}
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Tip dating and total evidence}

Two main features of new method for divergence-time dating:
\begin{itemize}
\item Explicit modelling of fossilization events as a part of the tree branching process (Pyron 2011, Heath {\it et al} 2014),
\item Utilising all existing data \textcolor{newblue1}{(total evidence)} in a joint inference (Pyron 2011, Ronquist {\it et al} 2012).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{frame}{Total evidence}

%\begin{itemize}

%\item Pyron 2011 uses Lewis MK model to model evolution of morphological characters.  
%\item Ronquist et al 2012, Wood et al 2012. 
%\item They used Yule or Uniform tree prior. 
%\end{itemize}

%\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Birth-death-sampling models}

\begin{figure}
  \begin{center}
               \input{pics/SampledTree1}
            \end{center}

         \end{figure}    
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Birth-death-sampling models}

 \begin{figure}
            \begin{center}
               \input{pics/SampledTree2}
            \end{center}

         \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame} {Fossillized birth-death model (FBD)}

\begin{columns}[t]
\begin{column}{.5\textwidth}

Stadler 2010, Heath {\it et al} 2014. 
\vspace{0.2cm}

The process starts at time $t_{or}>0$ and ends at time zero (present time).  
  \vspace{0.01cm}
   \begin{itemize}
\item birth rate $\lambda$
\item death rate $\mu$
\item sampling rate $\psi$
\item sampling at present probability $\rho$ 
   \end{itemize}
\vspace{0.01cm}

   \end{column}
   \begin{column}{.5\textwidth}

            \begin{center}
               \input{pics/SATree2}
            \end{center}

         \begin{center}\emph{Sampled tree} \end{center}

   \end{column}    
\end{columns}

\vspace{0.2cm}

Model parameters: \color{newblue1} $\eta = (t_{or}, \lambda, \mu, \psi, \rho)$.

\color{black}
All the parameters are identifiable.

\vspace{1.5cm}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame} {Fossilized birth-death (FBD) skyline model}

Stadler \& K\"uhnert et al 2012, Gavryushkina {\it et al} 2014.

\vspace{0.2cm}

There are $k$ time intervals and parameters remain constants within the intervals but may vary from one interval to another

\begin{columns}[c]
\begin{column}{.5\textwidth}
\begin{itemize}
\item birth rates $\lambda_1, \ldots, \lambda_k$
\item death rates $\mu_1, \ldots, \mu_k$
\item sampling rates $\psi_1, \ldots, \psi_k$
\item sampling at time $t_k$ (present) probability $\rho$
\end{itemize}

\begin{center}
Model parameters:
\color{newblue1} $\eta = (t_{or}, \bar \lambda, \bar \mu, \bar \psi, \rho)$

\end{center}
  \end{column}
   \begin{column}{.5\textwidth}
\begin{center}
\input{pics/SkylineTree}
\end{center}
\begin{center}\emph{\small Sampled tree} \end{center}   
 \end{column}    
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\iffalse

\begin{frame}{Variation in the choice of the tree model parameters}

The $t_{or}$ parameter can be replaced with $t_{root}$ parameter assuming the process started with a branching event.

\vskip2mm

Instead of $\lambda$, $\mu$, and $\psi$ we can also use an alternative parameterisation:

\renewcommand{\arraystretch}{1.2}

\begin{equation*} 
\begin{tabular}{ll} 
\text{net diversification rate} & $d = \lambda-\mu$  \\ 
\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
\end{tabular} \end{equation*}

\fi
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Heath {\it et al} approach and its extensions}

 Heath {\it et al} 2014 first used FBD model to infer divergence times of bears in a Bayesian MCMC framework:

\begin{itemize}
\item The topology of extant species is fixed
\item Only used occurrence dates of fossils, and no morphology 
\item Fixed $\rho$ to the truth in the inference
\end{itemize}

We extend this model in two ways:

\begin{itemize}
\item Sampling sampled ancestor trees (Gavryushkina {\it et al} 2014)
\item Incorparating morphological data
\end{itemize}

Implemented in SA and Morph-models packages for BEAST2 (\url{www.beast2.org})

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Comparative data of fossil tips vs only occurrence dates}

\begin{figure}
\includegraphics[width=\textwidth]{pics/Height_SA/Rplots.pdf}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Comparative data of fossil tips vs only occurrence dates}

\setlength{\oddsidemargin}{0.1in} % adjust as necessary

\begin{columns}[c]

\begin{column}{.7\textwidth}
\begin{center}
\begin{figure}
\includegraphics[height=0.65\textheight]{pics/TurnoverHPD/Rplots.pdf}
\end{figure}
\end{center}
 \end{column}

\begin{column}{.3\textwidth}
\begin{center}
turnover~rate~=~$\frac \mu \lambda$ 
\end{center}
 \end{column}    
\end{columns}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Biased estimates when not modelling sampled ancestors}

\begin{figure}
\includegraphics[width=\textwidth]{pics/SAvsNoSA/Rplots.pdf}
\end{figure}

\begin{center}
diversification rate = $\lambda-\mu$ 
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Analysis of penguin morphological data}

Penguin dataset (Ksepka et al 2012) consisting of morphological data of:
\begin{itemize}
\item all extant penguins (19 species)
\item 36 fossil species assigned to stratigraphic intervals
\end{itemize}

\vskip2mm

Models:

\begin{itemize}
\item Lewis MK vs MKv (Lewis 2001)
\item Partitions vs single alignment
\item Rate variation across sites and across partitions
\item Tree prior: FBD and Skyline FBD
\item Different tree model parameterisations
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}{Parameterisations}

\renewcommand{\arraystretch}{1.2}

lmp:

\color{newblue1}

\begin{equation*} 
\begin{tabular}{ll} 
\text{birth rate} & $\lambda$  \\ 
\text{death rate} & $\mu$ \\ 
\text{sampling rate} & $\psi$  \\ 
\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
\end{tabular} \end{equation*}

\textcolor{black}{dvs:}

\begin{equation*} 
\begin{tabular}{ll} 
\text{net diversification rate} & $d = \lambda-\mu$  \\ 
\text{turnover rate} & $\turnover = \frac \mu \lambda$ \\ 
\text{\sampleprop} & $\fosp = \frac\psi {\mu + \psi}$  \\ 
\textcolor{black}{other parameters} & \textcolor{black}{$t_{or}, \rho$} \\
\end{tabular} \end{equation*}

@@

\begin{frame}{But what about those hyper parameter priors?}
\begin{figure}
\includegraphics[width=0.6\textwidth]{priors.pdf}
\label{Fig:Prior}
\end{figure}

\small{Prior density for sampling times under fossilized birth-death process. Dot-dashed line uses priors from \textcite{gavryushkina2015bayesian}. 
Solid line is new prior with implicit assumptions on $T$ and $s$, dashed line results from only assuming implicit prior on $T$,  dotted line results from only assuming implicit prior on $s$.}
\end{frame}

\begin{frame}{And what about this guy?}
\includegraphics[width=0.5\textwidth]{Homo_naledi.jpg}
\includegraphics[width=0.5\textwidth]{Atlantic_Homo_naledi.jpg}

\bigskip{}

{\bf Can we date fossil remains using Bayesian phylogenetics? }
